Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-08-07 extended_catalog/paper/extended_catalog.tex
markus comments

However, a number of sources positionally coincident
with LAT sources exhibit extensiosn at other
wavelengths which are larger or comparable to the LAT PSF.
In particular certain SNRs, pulsar wind
nebulae (PWNe), molecular clouds, normal galaxies, 
galaxy clusters or dark
matter satellites can therefore be expected
to be observed as spatially extended sources in the LAT.
The LAT collaboration has previously reported on
the discovered of five spatially extended supernova remnants, IC443,
W28, W44, W51C, and RX J1713.7-3946
(\cite{ic443,w28,w44,w51c,rx_j1713_lat}). In addition, three extended
PWNe were detected: MSH 15-52, Vela~X, and
HESS\,J1825-137 (\cite{msh1552,velax,fermi_hess_j1825}), as well as
two close-by galaxies, the Large Magellanic and Small Magalenic Cloud
(\cite{lmc,smc}), and finally the radio galaxy Centarus A
(\cite{cen_a_lat}) .

@@

This analysis is not feasible using the official LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center (\cite{fssc})} because \gtlike can only 
fit the spectral parameters of the model. The extension fitting would
have to be done as an external loop which would typically be
too time consuming. What has typically
been done to study an extended sources is to fix the position of the
source and develop a profile of the likelihood as a
function of extension.  This approach is not optimal because the
position and extension of the source should be simultaneously maximized
to find the best fit parameters and to correctly compute the statistical
significance of the detection.  Furthermore since this approach is
computationally intensive, no large-scale Monte Carlo effort can
be run to validate it.

We developed an alternative approach to fit spatially extended
sources. A separate maximum likelihood fitting package called
\pointlike has been developed by the LAT collaboration and we have
added to it the ability to analyze spatially extended sources. \pointlike
is described in 1FGL and in Matthew Kerr's Ph.D thesis
(\cite{first_cat,matthew_kerr_thesis}).

What differentiates \pointlike most significantly from \gtlike is its
emphasis on speed by introducing approximations into the calculation
of the likelihood function while trying to minimized the introduced
numerical errors.  Like the binned version of \gtlike, \pointlike bins
the sky in both position and energy.  Unlike \gtlike, \pointlike relies on
the {\em healpix} representation of spatial bins (\cite{healpix}). Since
the PSF changes significantly with energy, the spatial size of each bin
is chosen to be smaller than the PSF. By changing the spatial
bin size with energy, \pointlike ensures that little information is lost
at each energy while minimizing the required number of bins.

\pointlike also bins $\gamma$-rays depending upon whether they convert
in the thin or thick layers of the tracker. Since the PSF is smaller for
photons which convert in the thin layers of the tracker,
photons are also binned depending upon where they 
convert in the tracker.
Furthermore, \pointlike
uses a sparse matrix representation of the spatial bins where only the
bins with counts in them are used. To evaluate the likelihood, \pointlike
then independently integrates the model predicted counts in the region.

Intelligent binning and a sparse matrix representation of the data
effectively interpolates between a binned analysis at low energy (where
the resolution is poor and there are many counts) and an unbinned
analysis at high energy (where the resolution is good but there are
few counts). This approach provides significant savings in time while
introducing only small numerical error.

>>>>

However, a number of sources positionally coincident
with LAT sources exhibit extension at other
wavelengths which are larger or comparable to the LAT PSF.
In particular certain SNRs, pulsar wind
nebulae (PWNe), molecular clouds, normal galaxies, 
galaxy clusters or dark
matter satellites can therefore be expected
to be observed as spatially extended sources in the LAT.
The LAT collaboration has previously reported on
the discovered of five spatially extended supernova remnants, IC443,
W28, W44, W51C, and RX J1713.7-3946
(\cite{ic443,w28,w44,w51c,rx_j1713_lat}). In addition, three extended
PWNe were detected: MSH 15-52, Vela~X, and
HESS\,J1825-137 (\cite{msh1552,velax,fermi_hess_j1825}), as well as
two close-by galaxies, the Large Magellanic and Small Magalenic Cloud
(\cite{lmc,smc}), and finally the radio galaxy Centarus A
(\cite{cen_a_lat}) .

@@

This analysis is not feasible using the standard LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center (\cite{fssc})} because \gtlike can only 
fit the spectral parameters of the model.

We note that \gtlike has been used in the past in several studies of
source extension in the LAT collaboration (\cite{lmc,smc,w28,w51c}).
Based upon a set of \gtlike maximum likelihood fits at fixed extensions,
a profile of the likelihood as a function of the extension was calculated.
This approach is not optimal because the position, extension, and
spectrum of the source must be simultaneously fit to find the best fit
parameters and to correctly compute the statistical significance of the
detection.  Furthermore since the \gtlike likelihood profile approach
is computationally intensive, no large-scale Monte Carlo effort can be
run to validate it.

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT collaboration and called \pointlike
(\cite{first_cat,matthew_kerr_thesis}).  We extended the code to allow
a simultaneous fit of the source extension together with the position
and the spectral parameters.

The choice to base the spatial extension fitting on \pointlike
rather than \gtlike was made on considerations of computing time.
The \pointlike algorithm was optimized for speed to handle
larger numbers of sources efficiently which is important for
our catalog scan as well as the Monte Carlo validation of our method.
Details on the \pointlike package can be found in (\cite{matthew_kerr_thesis}).
